 

### 2021-07-21 vertical-thermal-review.Rmd
small spelling edits

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021] to open forests and savannas [@curtis_intracanopy_2019; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated $CO_{2}$ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Elevated $CO_{2}$ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

@@

**Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

@@

**Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and LAI [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021]. Closed canopies, and high LAI canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate. The intensity of incoming radiation and VPD in the upper canopy can limit $g_s$. Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021]. In contrast, open canopies, or lower canopy LAI allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].

@@

It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems; for instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018]. 
Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig. `r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories). 
Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021].

@@

Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014]. Within-canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity,  @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.

@@

The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter biophysical and $T_{leaf}$ gradients in the forest [@rey-sanchez_spatial_2016].
Deciduousness in tropical forests is an adaptation to conserve water that is more common is dry than wet tropical forests. 
Within tropical deciduous species, the fraction of deciduous individuals tends to increase across tree size classes, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. 
In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons (e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018]. In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater $g_s$ thermal sensitivity of upper canopy sun leaves [@carterExperimentalWarmingTropical 2021]. Higher maximum temperatures, and $g_s$ influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.

>>>>

**Wind speeds are also higher at the top of the canopy, owing to the effect of buffering with canopy height [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @mcgregor_tree_nodate; @juckerCanopyStructureTopography2018] to open forests and savannas [@curtis_intracanopy_2019; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018], including all 22 forested NEON sites (Figs. `r fig_NEON_vertical`e, `r SIfig_NEON_all`). 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to 4.1 m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds averaged $\le$ 1.1 m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.**
Atmospheric CO~2~ concentrations tend to be higher near ground level at night, associated with plant and soil respiration, but the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of day when photosynthesis is occurring, pattern in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

@@

**Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar radiation loads, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @konradLeafTemperatureIts2021].
With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts2021; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

@@

**Vertical $T_{leaf}$ gradients can also vary with canopy structure (Fig. `r fig_leaf_T`).**
$T_{leaf}$ generally tracks $T_{air}$, where vertical gradients of $T_{air}$ depend on canopy structure and LAI [@leuzingerTreeSpeciesDiversity2007;@mullerEvidenceEfficientNonevaporative2021]. Closed canopies, and high LAI canopies act as an umbrella, absorbing most of the incoming radiation and preventing vertical air mixing in the understory microclimate. The intensity of incoming radiation and VPD in the upper canopy can limit $g_s$. Therefore, we would expect a stronger $T_{leaf}$ gradient where upper canopy leaves can reach much warmer temperatures than below canopies layers [@millerOnlySunlitLeaves2021]. In contrast, open canopies, or lower canopy LAI allow more vertical air mixing and sunlight into the understory, that can either neutralize a $T_{leaf}$ gradient or elevate below canopy $T_{leaf}-T_{air}$ of shade leaves more than upper canopy as in the case of the 'canopy greenhouse effect' or sunflecks, as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013b], and also due to the propensity of shade leaves for lower $g_s$ and greater thermal capacitance (capacity to hold heat) [@schymanskiStomatalControlLeaf2013b].

@@

It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems; for instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018; @stillImagingCanopyTemperature2021]. 
Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig. `r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories). 
Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature2021].

@@

Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014]. Within-canopy leaf nitrogen distribution can be regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. Therefore, we would expect elemental traits to influence vertical foliar thermal sensitivity,  however, @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.

@@

The timing of seasonal leaf loss or gain is influenced by vertical microclimate gradients, and in turn can alter biophysical and $T_{leaf}$ gradients in the forest [@rey-sanchez_spatial_2016].
Deciduousness in tropical forests is an adaptation to conserve water that is more common in dry than wet tropical forests. 
Within tropical deciduous species, the fraction of deciduous individuals tends to increase across tree size classes, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. 
In temperate deciduous forests, leaf-emergence is delayed with height-ontogeny, with understory species leafing out earlier than canopy trees [@vitasseEarlierLeafoutRather2014; @seiwaChangesLeafPhenology1999].

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and $T_{opt}$ across sites and seasons (e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a), one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5) in [Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of the RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) didnot differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018]. In contrast, tropical experimental warming observations show that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from understory to canopy, potentially linked to greater thermal sensitivity of $g_s$ in the upper canopy [@carterExperimentalWarmingTropical 2021]. Higher maximum temperatures, and $g_s$ are known influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear.
